Eco-evolutionary dynamics of finite populations from first principles

### 2023-02-27 ind_files/04_field_eqns.tex
Cite Fisher. Change adaptive dynamics to gradient dynamics

Adaptive dynamics is recovered under the following additional assumptions:

@@

Note that by definition, $B(y(t))$ is the variance of the trait in the population at time $t$. The term $w\left(z|N_K\delta_{y(t)}\right)$ is the expected growth rate of an individual with trait value $z$ in a population of size $N_K$ in which (almost) every individual has trait value $y$. This quantity is referred to as the `invasion fitness' of a `mutant' trait $z$ in a population of `resident' $y$ individuals. Comparing with \eqref{AD_canonical_eqn}, we can see that we have thus derived the canonical equation of adaptive dynamics. Note that strictly speaking, if $\psi(x,t) = \delta_{y(t)}$ exactly, then $B(y) \equiv 0$. This just reflects our assumption that mutations are sampled from infinitesimally close to the resident value. More detailed mathematical arguments are required to ensure that this convergence `makes sense' and that $B(y)$ does not actually equal 0. This has been proved using much more sophisticated mathematical tools in \citep{champagnat_unifying_2006}, which is where we refer the interested reader. An alternative heuristic argument is also provided in the classic article by \citep{dieckmann_dynamical_1996}.

>>>>

Some more familiar dynamics are recovered under the following additional assumptions:

@@

Note that by definition, $B(y(t))$ is the variance of the trait in the population at time $t$. The term $w\left(z|N_K\delta_{y(t)}\right)$ is the expected growth rate of an individual with trait value $z$ in a population of size $N_K$ in which (almost) every individual has trait value $y$. This quantity is referred to as the `invasion fitness' of a `mutant' trait $z$ in a population of `resident' $y$ individuals. Equation \eqref{AD_canonical_derived} represents a broad class of systems called `gradient equations' in quantitative genetics \citep{lande_natural_1976, abrams_relationship_1993, lehtonen_price_2018, lion_theoretical_2018}, and captures the approximate evolutionary dynamics of quantitative traits under certain mutation limits.  Comparing with \eqref{AD_canonical_eqn}, we can see that this is very similar to the canonical equation of adaptive dynamics, and indeed, the two are deeply related \citep{lehtonen_price_2018, lion_theoretical_2018}. The major difference is that the `proper' canonical equation of adaptive dynamics as formulated in \cite{dieckmann_dynamical_1996}, the function $B(y)$ explicitly relies on mutations as a continual source of variation, whereas in gradient dynamics and our equation \eqref{AD_canonical_derived}, $B(y)$ captures the standing genetic variation in the population but does not specify the source of this variation. Note that strictly speaking, if $\psi(x,t) = \delta_{y(t)}$ exactly, then $B(y) \equiv 0$. This just reflects our assumption that mutations are vanishingly rare and mutants are sampled from infinitesimally close to the resident value. More detailed mathematical arguments are required to ensure that this convergence `makes sense' and that $B(y)$ does not actually equal 0. This has been proved rigorously using much more sophisticated mathematical tools grounded in martingale theory \citep{champagnat_unifying_2006}. A similar heuristic derivation of the canonical equation of adaptive dynamics is also provided in the classic article by \cite{dieckmann_dynamical_1996}.

### 2023-02-27 ind_files/06_unification_discussion.tex
Cite Fisher. Change adaptive dynamics to gradient dynamics

        $m = \infty$ \newline (Quantitative traits) & $\left\{\sum\limits_{i=1}^{n}\delta_{x_i} \ | \ n \in \mathbb{N}\right\}$ \newline \newline (Each Dirac mass $\delta_{x_i}$ is an individual with trait value $x_i$) & Two real-valued functionals $b(x|\nu)$ and $d(x|\nu)$ describing the birth and death rate of trait variant $x$ when the population is $\nu$ & Functional Fokker-Planck equation/Field theory \newline (SPDEs) & Kimura's continuum-of-alleles model \newline \cite{sasaki_oligomorphic_2011}'s Oligomorphic Dynamics \newline \cite{wickman_theoretical_2022}'s Trait Space Equations for intraspecific trait variation \newline Adaptive Dynamics  \newline Price equation (quantitative traits)\\

@@

Taking $K \to \infty$ in \eqref{nD_stochastic_Fisher} recovers a well-known equation in population genetics upon noting that the process tends to a deterministic process as $K \to \infty$, as noted in section \ref{sec_nD_det_limit}, and thus the expectation value in the infinite population case is superfluous. The first term, $\sigma^2_w$, is the subject of Fisher's fundamental theorem \citep{price_fishers_1972, frank_fishers_1992, kokko_stagnation_2021}. The last term arises in both finite and infinite populations whenever $w_i$ can vary over time \citep{baez_fundamental_2021}, be it through frequency-dependent selection, phenotypic plasticity, varying environments, or other ecological mechanisms, and represents feedback effects on the fitness $w_i$ of the $i\textsuperscript{th}$ species over short (`ecological') time-scales. The fact that Fisher appears to have been rather dismissive of this feedback, often only referring to it in a cursory manner as `environmental change' (or `environmental degradation'), has led to much discussion, debate, and confusion about the interpretation, importance, and implications of his `fundamental theorem' (see \cite{kokko_stagnation_2021} and sources cited therein). Finally, the second term of equation \eqref{nD_stochastic_Fisher} is a manifestation of noise-induced selection and vanishes in the infinite population limit, and is thus particular to finite populations.\\

>>>>

        $m = \infty$ \newline (Quantitative traits) & $\left\{\sum\limits_{i=1}^{n}\delta_{x_i} \ | \ n \in \mathbb{N}\right\}$ \newline \newline (Each Dirac mass $\delta_{x_i}$ is an individual with trait value $x_i$) & Two real-valued functionals $b(x|\nu)$ and $d(x|\nu)$ describing the birth and death rate of trait variant $x$ when the population is $\nu$ & Functional Fokker-Planck equation/Field theory \newline (SPDEs) & Kimura's continuum-of-alleles model \newline \cite{sasaki_oligomorphic_2011}'s Oligomorphic Dynamics \newline \cite{wickman_theoretical_2022}'s Trait Space Equations for intraspecific trait variation \newline Gradient Dynamics \newline Price equation (quantitative traits)\\

@@

Taking $K \to \infty$ in \eqref{nD_stochastic_Fisher} recovers a well-known equation in population genetics upon noting that the process tends to a deterministic process as $K \to \infty$, as noted in section \ref{sec_nD_det_limit}, and thus the expectation value in the infinite population case is superfluous.The first term, $\sigma^2_w$, is the subject of Fisher's fundamental theorem \citep{fisher_genetical_1930,  price_fishers_1972, frank_fishers_1992, kokko_stagnation_2021}. The last term arises in both finite and infinite populations whenever $w_i$ can vary over time \citep{baez_fundamental_2021}, be it through frequency-dependent selection, phenotypic plasticity, varying environments, or other ecological mechanisms, and represents feedback effects on the fitness $w_i$ of the $i\textsuperscript{th}$ species over short (`ecological') time-scales. Fisher appears to have been rather vague and dismissive of this feedback \citep{fisher_genetical_1930}, and this has led to much discussion, debate, and confusion about the interpretation, importance, and implications of his `fundamental theorem' (see \cite{kokko_stagnation_2021} and sources cited therein). Finally, the second term of equation \eqref{nD_stochastic_Fisher} is a manifestation of noise-induced selection and vanishes in the infinite population limit, and is thus particular to finite populations.\\
